An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-06-29 main.tex
Dagens text

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=true}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

\usepackage{amsthm}
\usepackage{amsmath,mathtools,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

Many other generalisations of the Parikh map have been studied, most of them
less plain than this one. Prominent examples include generalising the Parikh map
to segments of a fixed length \cite{KARHUMAKI1980155} and the more general
Parikh matrix, which gives more information about a word than the standard
Parikh image \cite{parikh-matrix}. Another notable generalisation is the
p-vector, introduced in~\cite{infinite-words}, which denotes the position of
each letter in the word rather than the number of their occurrences and allows
for generalisations into infinite alphabets. All of these, in some sense, add
something which is not already there in the Parikh map. By contrast, the main
utility of the formulation introduced here is that it allows us to \emph{remove}
something, thereby obtaining an easier problem.

@@

\begin{definition}
  The Parikh predicate, $\SinglePredicateInstance$, for some automaton
  $\Automaton = \AutomatonTuple$, modulo some map $\Map$ to a commutative monoid
  $\Monoid$ as described in Section~\ref{sec:generalised} and with a transition
  selection function $\Filter:\: \Transitions \to \Naturals$ holds when
  $\MonoidElement \in \Monoid$ is the Parikh image of $\Automaton$ modulo
  $\Map$, or more formally when there is an accepting path $\Path =
  \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such that:
  \begin{itemize}
    \item $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
    \item $\MonoidElement = \Map(\WordOf(\Path))$
  \end{itemize}
\end{definition}

@@

\subsection{The calculus is correct}

To be as clear as possible about what it means for our calculus to be correct,
we use more exact definitions than the traditional soundness and completeness.

\subsubsection{The calculus terminates}
\subsubsection{The calculus allows no values outside the Parikh image}
\subsubsection{The calculus excludes no values in the Parikh image}

\subsection{Finding the Presburger representation of a Parikh Image}
\contents{
  \item lazy qe
}

@@

To generalise the calculus to calculations on products of automata, we change the main predicate to take arbitrarily many automata:
\begin{definition}
  $\ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}$
  is true exactly when the single-automaton version of the predicate would hold
  for the automaton.
\end{definition}

@@

\subsection{The extended calculus is also correct}
\subsubsection{The calculus terminates}
\subsubsection{The calculus allows no values outside the Parikh image}
\subsubsection{The calculus excludes no values in the Parikh image}

@@

\Catra{} is written in Scala, with the calculus described in this paper
implemented as a theory plug-in for the \Princess{} automated theorem
prover~\cite{princess}, which also performs the Presburger reasoning. For
comparison, we also provide an implementation of the baseline method
from~\cite{generate-parikh-image}, a direct translation that uses the~\Nuxmv{}
symbolic model checker~\cite{nuxmv} to solve our constraints, and \Fudge{the
approximation described in~\cite{approximate-parikh} on top of the standard
baseline back-end}. An example of an input file \Fudge{illustrating every feature
of the file format} can be seen in Listing~\ref{lst:input-example}.

@@

%% Appendix
\appendix
% \section{Appendix}

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous,screen]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=true}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}

\usepackage{mathtools} % Bonus

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

Many other generalisations of the Parikh map have been studied, most of them
less plain than this one. Prominent examples include generalising the Parikh map
to segments of a fixed length \cite{KARHUMAKI1980155} and the more general
Parikh matrix, which gives more information about a word than the standard
Parikh image \cite{parikh-matrix}. Another notable generalisation is the
p-vector, introduced in~\cite{infinite-words}, which denotes the position of
each letter in the word rather than the number of their occurrences and allows
for generalisations into infinite alphabets. All of these, in some sense, add
something which is not already there in the Parikh map. By contrast, the main
utility of the formulation introduced here is that it allows us to \emph{remove}
something, thereby potentialy obtaining an easier problem.

@@

\begin{definition}\label{def:single-image}
  The Parikh predicate, $\SinglePredicateInstance$, for some automaton
  $\Automaton = \AutomatonTuple$, modulo some map $\Map$ to a commutative monoid
  $\Monoid$ as described in Section~\ref{sec:generalised} and with a transition
  selection function $\Filter:\: \Transitions \to \Naturals$ holds when
  $\MonoidElement \in \Monoid$ is the Parikh image of $\Automaton$ modulo
  $\Map$, or more formally when there is an accepting path $\Path =
  \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such that:
  \begin{itemize}
    \item $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
    \item $\MonoidElement = \Map(\WordOf(\Path))$
  \end{itemize}
\end{definition}

@@

\subsection{\Calculus{} for one automaton is correct}\label{sec:single:correct}

To be as clear as possible about what it means for our calculus to be correct,
we use more exact definitions than the traditional soundness and completeness.
Since correctness is guaranteed by the formulation of the predicate, our
correctness proof means proving inductively that no rule would invalidate that
definition.

\subsubsection{\Calculus{} terminates}
\begin{lemma}\label{lma:single-terminates}
  A solver implementing the calculus described in \cref{sec:calculus} starting
  with some decidable (modulo the solver) set of clauses $\SomeClause{}$ and an
  instance of our predicate $\SinglePredicateInstance$ will eventually
  terminate.
\end{lemma}

\begin{proof}
The first remaining source of divergence is from leftover predicates. The
$\Image$ predicate can always be trivially removed by applying \Expand{}. This
means a leftover $\Image$ predicate can never cause divergence. The only
remaining predicate then is $\Connected$, which can always be removed by
applying $\Subsume$ when no other of the rules is applicable. Hence all
predicates introduced in the calculus can always be removed.

It remains to prove that each rule is only applied finitely many times. This is
almost obvious since both \Propagate{} and \Split{} can only be applied to
transition variables whose $\Filter$ value has no known bound. However, \Split{}
can always be used to split the proof tree and add clauses in precisely the
right shape, thereby guaranteeing the ability to make forward progress.
Therefore, the number of transition variables whose bounds are unknown is
decreasing monotonically for any given proof goal, and hence termination is
guaranteed.
\end{proof}

\subsubsection{\Calculus{} computes the homomorphic image}
\begin{lemma}\label{lma:single-correct}
  Starting with $\SinglePredicateInstance$, we will close the goal with $\top$
  if $\exists s \in \Language(\Automaton) \HoldsThat \Map(s) = \MonoidElement$
  and $\bot$ otherwise.
\end{lemma}
\begin{proof}
Since we are always guaranteed to terminate following
\cref{lma:single-terminates}, we know that we will close the goal eventually.
Additionally, the property of \cref{lma:single-terminates} is trivially obtained
from the semantics of the $\Image$ predicate as defined in
\cref{def:single-image}. This means that we only need to prove the preservation
of those properties to prove \cref{lma:single-correct}.

\begin{lemma}\label{lma:expand-preserves}
  \Expand{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}

  After application of \Expand{}, $\FlowEq$ guarantees that any
  $\Filter(\Transition)$  for every $\Transition \in \Automaton$ is consistent
  with the flow through the automaton. After its application, the proof goal
  either contains an unsatisfiable set of Presburger formulae if the flow is
  infeasible, otherwise a satisfiable set. An infesaible flow can happen in
  either of these conditions:
  \begin{itemize}
    \item if $\Filter$ demands that some transition $\Transition$ is visited more times than would be allowed by $\Automaton$, or
    \item if the path described by $\Filter$ is not consistent with a path from
    $\InitialState$ to some state $\State$ in $\AcceptingStates$.
  \end{itemize}

  Note that \Expand{} is taken directly from the existing formulation of
  \cref{eq:generate-parikh}.
\end{proof}

\begin{lemma}\label{lma:split-preserves}
  \Split{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}
  \Split{} allows us to case-split at most once on each transition $\Transition
  \in \Transitions$. This means that our proof tree at its leaves will be
  guaranteed to have all transitions of known bounds in the correct format. This
  means that once \Split{} is at fixpoint, the status (present or absent) of any
  transition is known. Therefore it preserves the semantics of the $\Image$
  predicate.
\end{proof}

\begin{lemma}\label{lma:propagate-preserves}
  \Propagate{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}
  \Propagate{} will lazily enforce connectedness to any state. If it can be
  applied, there exists a cut $C$ in $\Automaton$ such that some state $\State$
  for any incoming transition $\Transition \in \Transitions_\Automaton$ has
  $\Filter(t) = 0$  as part of the equations of the proof goal, and such that
  $\State$ has an outgoing transition $\Transition'$ whose lower bound is not
  $0$. Adding this conclusion will make equations unsatisfiable precisely if
  there exists a transition $\Transition \in \Transitions_\Automaton$ such that
  $\Filter(\Transition)$ is inconsistent with the homomorphic image, since it
  implies that the selection of transitions is disconnected.
\end{proof}

\begin{lemma}\label{lma:subsume-preserves}
  \Subsume{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}
  Since this rule can \emph{only} be applied if no other rule can, it means that
  the upper and lower bounds of any transition $\Transition \in \Transitions$ as
  defined by $\Filter(\Transition)$ is known, meaning clauses of the shape
  $\Filter(\Transition) > 0$ or $\Filter(\Transition) = 0$ are present in the
  proof goal for every transition and that the $\Image$ predicate has been
  consumed by \Expand{} to be replaced by $\Connected$. This in turns means that
  the goal can be closed only if the assignment of $\Filter$ is consistent with
  an accepting path through $\Automaton$.

  It also means that \Propagate{} must have been applied whenever possible, meaning that for any transition $\Transition$ starting in a state $\State$ that is disconnected the corresponding $\Filter$ values, we must have applied $\Propagate{}$ to derive $\Filter(\Transition) = 0$. Note that for an infeasible assignment (such as an unsatisfiable instance) this would contradict other decisions, but it would preserve the semantics of $\Image$.
\end{proof}

Since the final step we execute is \Subsume, which we know from
\cref{lma:subsume-preserves} preserves the semantics of $\Image$ and each step
along the way
(\cref{lma:expand-preserves,lma:split-preserves,lma:propagate-preserves}), it
holds that \Calculus{} for one automaton also preserves the semantics.
\end{proof}

\subsection{Finding the Presburger representation of a homomorphic image}
\contents{
  \item lazy qe
}

@@

To generalise the calculus to calculations on products of automata, we change the main predicate to take arbitrarily many automata:
\begin{definition}\label{def:multiple}
  $\ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}$
  is true exactly when the single-automaton version of the predicate would hold
  for the automaton.
\end{definition}

@@

\subsection{The extended calculus is also correct}

Since the expanded calculus is in practice the addition of two rules with the
purpose of reducing an~$\Image$ predicate containing a product to the form
solved by the previous single-automaton rules of \cref{tbl:rules:single}, we
extend the reasoning from \cref{sec:single:correct} to the single-automaton
rules of \cref{tbl:rules:multi}.

\subsubsection{The calculus terminates}
\begin{lemma}
  The addition of the \ExpandM{} and \Materialise{} rules in \cref{tbl:rules:multi}
  maintains termination as shown in \cref{lma:single-terminates}.
\end{lemma}
\begin{proof}
  Similarly to how we previously showed that the number of possible executions
  of the \Split{} rule is bounded above by the number of transitions of an
  automaton, we can bound the number of applications of \Materialise{} by the
  (monotonically) decreasing number of automata in the product until we have
  approached the starting state for the single-automaton calculus. Trivially, we
  can eagerly apply \Materialise{} repeatedly to do so. Similarly, \ExpandM{} is
  just the generalisation of \Expand{} to multiple automata. In other words there are no possibilities of the calculus diverging.
\end{proof}

\subsubsection{The calculus implemements the homomorphic image of the product}
\begin{lemma}
  Using \Calculus{} to compute $\Image_{\Automaton_1\times\ldots\times\Automaton_k, \Map}(\Filter, \MonoidElement)$ is equivalent to using the single-automaton rules to compute $\Image_{\Automaton', \Map}(\Filter, \MonoidElement)$ where $\Automaton' = \Automaton_1\times\ldots\times\Automaton_k$.
\end{lemma}
\begin{proof}

  We perform this proof equivalent to set inclusion: by proving that neither predicate is stronger than the other, in the sense of rejecting an element that the other accepts and vice versa.

\begin{lemma}\label{lma:multi:rinclude}
  There exists no element $\MonoidElement$ such that $\Image_{\Automaton_1\times\ldots\times\Automaton_k, \Map}(\Filter, \MonoidElement) = \top$ but $\Image_{\Automaton', \Map}(\Filter, \MonoidElement) = \bot$.
\end{lemma}
\begin{proof}
  Assume that such an $\MonoidElement$ exists with the goal of deriving a
  contradiction. Under this assumption, would we be able to use the two new
  rules to close the proof goal with $\top$?

  The \ExpandM{} rule would allow us to close no goal since it only adds
  additional clauses. Our only hope is in using \Materialise{} to get rid of the
  product step by step without introducing an unsatisfiable clause. However, all
  \Materialise{} does is computing the product step by step, eventually arriving
  at $\Automaton'$ modulo permutations. Therefore, it can be no more satisfiable
  than the single-automata predicate.

  \contents{\item Do I need to worry about the fact that I have no extra
  clauses/$\SomeClause$ here? If present, I would need to apply the definition
  of \BindingSum{} since that is what transfers constraints on individual
  automata up and down the materialisation.}
\end{proof}

\begin{lemma}\label{lma:multi:linclude}
  There exists no element $\MonoidElement$ such that $\Image_{\Automaton_1\times\ldots\times\Automaton_k, \Map}(\Filter, \MonoidElement) = \top$ but $\Image_{\Automaton', \Map}(\Filter, \MonoidElement) = \bot$.
\end{lemma}
\begin{proof}
  Assume that such an element exists. Then $\MonoidElement =
  \Map(\WordOf(\Path))$ for some $\Path$ ending up in an accepting state in
  $\Automaton'$, and consistent with $\Filter$ along the transitions of $\Path$.
  Then a corresponding path must exist in each automata of $\Automaton'$,
  although possibly with different labels.

  For \ExpandM{} to introduce unsatisfiability, either of its automata must be
  disconnected (or its corresponding $\Connected{}$ predicate can be removed
  with \Propagate{}), or its flow must be infeasible. However, this contradicts
  the existence of some corresponding path $\Path$ in each automaton. Hence,
  \ExpandM{} cannot be the source of the unsatisfiability.

  The same applies for the second half of \Materialise{}, if the corresponding
  path exists in each individual automaton and in $\Automaton'$, it must also
  exist in every intermittent product on the way to $\Automaton'$, and if $\Map$
  is consistent for intermittent transition labels as it must be, then neither the intermittent $\Image$ predicates nor the final one can be the source of the unsatisfiability.

  This leaves one final potential source of unsatisfiability: the first half of
  \Materialise{}: \BindingSum{}. For that to be unsatisfiable, it requires
  intermittent automata $\Automaton_l, \Automaton_r$ such that adding
  $\BindingSum(\Automaton_l \times \Automaton_r)$ is unsatisfiable. The
  generated equations state that the transitions of $\Automaton_l$ and
  $\Automaton_r$ must be used precisely as often as their corresponding results
  in $\Automaton_l \times \Automaton_r$. For this set of equations to be
  unsatisfiable there must exist no path through $\Automaton_l \times
  \Automaton_r$ that preserves this. However, this is true from the definition
  of the product of automata: since we know the product to be nonempty, it must
  in some sense contain at least one path whose equivalent exists in both
  $\Automaton_l$ and $\Automaton_2$ and cannot use a corresponding transition
  more often, or it would recognise a different langauge. Hence no such case
  exists.
\end{proof}

Since it follows from \cref{lma:multi:linclude,lma:multi:rinclude} that both
variants of the predicate are precisely equally strong, it must mean that they
are equivalent, and we have proven the correctness as defined in
\cref{def:multiple}.
\end{proof}

@@

\Catra{} is written in Scala, with the calculus described in this paper
implemented as a theory plug-in for the \Princess{} automated theorem
prover~\cite{princess}, which also performs the Presburger reasoning. For
comparison, we also provide an implementation of the baseline method
from~\cite{generate-parikh-image}, a direct translation that uses the~\Nuxmv{}
symbolic model checker~\cite{nuxmv} to solve our constraints, and \Fudge{the
approximation described in~\cite{approximate-parikh} on top of the standard
baseline back-end}. An example of an input file \Fudge{illustrating every feature
of the file format} can be seen in \cref{lst:input-example}.

@@

%% Appendix
\appendix
\section{Appendix}
